Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees

### 2023-01-14 script/manuscript_MEE.Rmd
some authorities and rearrangement of appendices

---
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$ 
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU
    |  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie$^3$$^,$$^4$ , Emily J. Woodworth$^1$$^,$$^2$  
abstract:  |  
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblages. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods. 
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.  
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.  
keywords: | 
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

We then generated 4,029 absence points , locations where the focal taxon is anticipated missing, through a random stratification of 19% of the land cover in the area and included them in (BLM CITATION - need appropriate format for journal). To achieve a larger absence data set, we generated 1,000 pseudo-absence records for each taxon by randomly selecting coordinates located at least 10km away from an occurrence record. For ML models, these pseudo-absences were reduced so that the ratio of presence to absence records were balanced (@barbet2012selecting). To achieve this, we removed absence records inside of 10% of the mean sample value of the presence records; the required number of absence records were then randomly sampled.

@@

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae less Heliantheae. Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`. Both morphotypes of Asteraceae pollen `r toString(dd)`, and Ericaceae `r toString(ee)`.

@@

```{r Correlation between Number of Grains and Number of Sequences, eval = F}

@@

ct <-cor.test(pol_p$countProp, pol_p$adjSeqProp, method = "kendall")
# ct[["p.value"]] too small to write out. tau > 0.4 strong relationship
summary(model <-  lm(adjSeqProp ~ countProp, data = pol_p) )

@@

rm(pol_p, newdat)
```

To detect whether the sequencing reads were semi-quantitative the proportion of all pollen morphotypes distinguishable by microscopy were compared to the sequence reads. In all instances sequence reads were pooled to the highest taxonomic rank associated with the morphotype, e.g. if both species of *Mertensia*, or one species and read only classified to genus were present in a sample, the reads were summed. The total percentage of the ten most abundant grains per sample were then were then *corrected* to constitute the entire sample.

@@

```{r plot phenology estimates, echo=F, fig.cap="Modelled dates of when major flowering events occurred", out.width = '75%'}
knitr::include_graphics("../graphics/plots/phenology_estimates_10_90.png")
# these are the tenth and nintey  percent non-parametric confidence intervals. 
```

@@

For example a common UNKNOWN sequence mapped to the Asteraceae family, but which was flagged by temporal filters and is present in both
*B. nevadensis* and *B. rufocinctus* pollen is most likely *Frasera*, failed extraction. A similar likely mismatch could be between what was fide molecular evidence as *Agastache pallidiflora* but where feeding was infrequently observed on *Pedicularis*, likely due to this entire order being represented by only a single molecular reference species.

@@

... many of our results indicate foraging on *Viola* spp, zygomorphic flowers with architecture which would require subtle handling and strength to reach the pollen and nectar loads... .. Or the *Epilobium sp.* results indicating that a species such as *Chamerion angustifolium* or *latifolium* is occasionally utilized, as it supported by limited paylnological data.

@@

We have demonstrated how Angiosperms533 hyb-seq probes may be used for plant barcoding in a metagenomic context. This was exemplified in an ecologically relevant scenario, where the results have immediate implications for natural history driven fundamental science and the applied science of land management. The test pollen loads contained a number of closely related taxa, some in notoriously morphologically difficult clades with rapid rates of diversification (e.g. *Mertensia*, *Lupinus*), at naturally occurring proportions (@nevado2016widespread, @nazaire2014phylogenetic).  We incorporated spatial and temporal approaches for creating custom sequence databases an approach which is readily applicable to any lab group with the capacity to perform next-generation sequencing across the entirety of multiple continents, and which we expect to be highly beneficial in many study areas. By combining insights from these novel approaches with an extensive observational field based study we show how these methods may be applied to test a variety of hypotheses related to ecological interactions.

We anticipate that many of the complications which we faced, using opportunistically collected pollen loads and the first implementation of this method may readily be overcome. It seems apparent that we had issues detecting pollen from several genera of plants, based upon these and other observational studies most likely *Vicia*, *Lathyrus*, and *Frasera* (@inouye1980effect, @pleasants1980competition), this is most likely related to user error in obtaining high quality DNA during the plant reference library generation period.  **(REED SHOULD HUNT FOR LOCI RETURNS FROM BLAST LIKE HE DID POLLEN)**. Additional complications seem to relate to the presence of closely related false positives, e.g. frequent classifications of sequences as *Trollius*,  *Caltha*, and *Thalictrum* alongside a more common species in the family, e.g. *Delphinium.*. Many of our errors are known to us and multiple mnemonics are in *APPENDIX XX* to assist others in future attempts to achieve better results.  However, the line between false and negative positives may be blurred in some of these instances and warrant further work, for example **Ericaceae pollen grains were observed in a number of samples in trace quantities...**.

@@

  The remarkably clear foraging preferences of *Bombus*, both here and across a great many localities globally, reiterates the needs for land managers to maintain relatively high amounts of members of the Fabaceae, Boraginaceae, and Ranunculaceae across the landscape. Numerous historic, and some ongoing, land management actions reduce the ability of many landscapes to support stable populations of *Bombus*. Historic, and occasionally current, livestock grazing may be associated with the targeted removal of many species of plants which are known to have compounds toxic to cattle. In particular, the removal of locoweeds (Fabaceae: *Astragalus* & *Oxytropis*) and larkspurs (Ranunculaeae: *Delphinium*) were common across public lands administered by the United States Forest Service, and to an extent the Bureau of Land Management (@ralphs1988herbicide, @aldous1919eradicating, @ralphs2003mechanism). Further actions, generally initiated by settlers to gain water rights or to support mining operations, involved the channelization and incising of streams, and culling of beavers, reducing the extent of wetlands and associated mesic meadows which provide habitat for many species of *Mertensia* (Boraginaceae) widely distributed across Western North America (). Fire suppression further resulted in the succession of many Aspen groves to Conifer stands, decreasing the mosaic of age structured habitats in many landscapes, adversely effection habtitat for tall *Mertensia* species. Finally the effects of Nitrogen deposition, especially given the West's growing population still pose adverse effects on the abundance of a variety of species of Fabaceae (see @stevens2018atmospheric). Current solutions to these issues, involve targeted burns, reintroduction of beavers and beaver habitats, and the possibility of re-seeding a variety of 'locoweeds' and 'larkspurs' in areas now seldom used for grazing. The highly enthusiastic response of land managers, and homeowners, to plant *Ascelpias* to improve Monarch Butterfly (*Danaus plexippus*) habitat provides an effective framework for the latter.

@@

Appendix XX - Origins of Corbiculae Loads

@@

@@

Appendix XX - Species Distribution Models Predictors

Layer                   Description                            Source
----    ----------------------------------------------  ---------------------
1.             Mean annual cloudiness - MODIS            Wilson et al. 2016
2.           Cloudiness seasonality 1 - MODIS            Wilson et al. 2016 
3.           Cloudiness seasonality 2 - MODIS            Wilson et al. 2016 
4.           Cloudiness seasonality 3 - MODIS            Wilson et al. 2016
5.          Beginning of the frost-free period               Wang et al.
6.             Climatic moisture deficit                     Wang et al.
7.             Degree-days above 5C from                     Wang et al.
8.             Mean annual precipitation                     Wang et al.
9.          Mean annual precipitation as snow                Wang et al.
10.             Temperature seasonality                      Wang et al.
11.       2015 Percent Grass/Herbaceous cover - MODIS         (MOD44B)
12.      2015 Percent Tree cover from Landsat 7/8              (GLCF)
13.       Soil probability of bedrock (R Horizon)             SoilGrids
14.         Soil organic carbon (Tonnes / ha)                 SoilGrids
15.              Surface soil pH in H2O                       SoilGrids
16.             Surface soil percent sand                     SoilGrids
17.                   Soil USDA class                         SoilGrids
18.                Topographic elevation                    EarthEnv DEM
19.          Topographic elevation, moving window.          EarthEnv DEM
20.            Topographic percent slope                    EarthEnv DEM 
21.            Topographic wetness index                    EarthEnv DEM 
22.             Topographic aspect from                     EarthEnv DEM
23.      Annual potential solar radiation computed             r.sun
24.       Estimated actual (w/-cloud) solar radiation  r.sun / Wilson et al. 2016
25.       Log-transformed distance to surface water   Global Surface Water Explorer
26.             Percent surface water                 Global Surface Water Explorer 
----    ----------------------------------------------  ---------------------

@@

Appendix XX - Molecular Reference Specimen Table

@@

Appendix XX - All Pollen Reference Slides Used to Establish Morphotypes

@@

Random Forest have high bias and low variance, where boosted regressions trees have low bias and high variances. 
\newpage

>>>>

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    |  Reed Clark Benkendorf$^1$$^,$$^2$, Emily J. Woodworth$^1$$^,$$^2$, Jane E. Ogilive$^3$$^,$$^4$,   
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
    |  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie$^3$$^,$$^4$ , Emily J. Woodworth$^1$$^,$$^2$  
abstract:  |  
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblages. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods.  
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.  
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.  
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

We then generated 4,029 absence points , locations where the focal taxon is anticipated missing, through a random stratification of 19% of the land cover in the area and included them in (@aim2019database). To achieve a larger absence data set, we generated 1,000 pseudo-absence records for each taxon by randomly selecting coordinates located at least 10km away from an occurrence record. For ML models, these pseudo-absences were reduced so that the ratio of presence to absence records were balanced (@barbet2012selecting). To achieve this, we removed absence records inside of 10% of the mean sample value of the presence records; the required number of absence records were then randomly sampled.

@@

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae Berchtold & J. Presl less Heliantheae. Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`. Both morphotypes of Asteraceae pollen `r toString(dd)`, and Ericaceae `r toString(ee)`.

@@

```{r Correlation between Number of Grains and Number of Sequences}

@@

ct <-cor.test(pol_p$countProp, pol_p$adjSeqProp, method = "kendall")
# ct[["p.value"]] too small to write out. tau > 0.4 strong relationship
model <-  lm(adjSeqProp ~ countProp, data = pol_p)

@@

rm( newdat)
```

To detect whether the sequencing reads were semi-quantitative the proportion of all pollen morphotypes distinguishable by microscopy were compared to the sequence reads. In all instances sequence reads were pooled to the highest taxonomic rank associated with the morphotype, e.g. if both species of *Mertensia* Roth, or one species and read only classified to genus were present in a sample, the reads were summed. The total percentage of the ten most abundant grains per sample were then were then *corrected* to constitute the entire sample.

@@

```{r plot phenology estimates, echo=F, fig.cap="Modelled dates of when major flowering events occurred", out.width = '100%'}
knitr::include_graphics("../graphics/plots/phenology_estimates_10_90.png")
# these are the tenth and nintey  percent non-parametric confidence intervals. 
```

@@

For example a common UNKNOWN sequence mapped to the Asteraceae family, but which was flagged by temporal filters and is present in both
*B. nevadensis* Cresson and *B. rufocinctus* Cresson pollen is most likely *Frasera* Walter, failed extraction. A similar likely mismatch could be between what was fide molecular evidence as *Agastache pallidiflora* (A. Heller) Rydb. but where feeding was infrequently observed on *Pedicularis* L., likely due to this entire order being represented by only a single molecular reference species.

@@

... many of our results indicate foraging on *Viola* L. spp, zygomorphic flowers with architecture which would require subtle handling and strength to reach the pollen and nectar loads... .. Or the *Epilobium* L. spp. results indicating that a species of *Chamerion* Seg. such as *C. angustifolium* (L.) Scop.  or *latifolium* (L.) Sweet is occasionally utilized, as it supported by limited paylnological data.

@@

We have demonstrated how Angiosperms533 hyb-seq probes may be used for plant barcoding in a metagenomic context. This was exemplified in an ecologically relevant scenario, where the results have immediate implications for natural history driven fundamental science and the applied science of land management. The test pollen loads contained a number of closely related taxa, some in notoriously morphologically difficult clades with rapid rates of diversification (e.g. *Mertensia*, *Lupinus* L.), at naturally occurring proportions (@nevado2016widespread, @nazaire2014phylogenetic).  We incorporated spatial and temporal approaches for creating custom sequence databases an approach which is readily applicable to any lab group with the capacity to perform next-generation sequencing across the entirety of multiple continents, and which we expect to be highly beneficial in many study areas. By combining insights from these novel approaches with an extensive observational field based study we show how these methods may be applied to test a variety of hypotheses related to ecological interactions.

We anticipate that many of the complications which we faced, using opportunistically collected pollen loads and the first implementation of this method may readily be overcome. It seems apparent that we had issues detecting pollen from several genera of plants, based upon these and other observational studies most likely *Vicia* L., *Lathyrus* L., and *Frasera* (@inouye1980effect, @pleasants1980competition), this is most likely related to user error in obtaining high quality DNA during the plant reference library generation period.  **(REED SHOULD HUNT FOR LOCI RETURNS FROM BLAST LIKE HE DID POLLEN)**. Additional complications seem to relate to the presence of closely related false positives, e.g. frequent classifications of sequences as *Trollius* L.,  *Caltha* L., and *Thalictrum* L. alongside a more common species in the family, e.g. *Delphinium.* L. Many of our errors are known to us and multiple mnemonics are in *APPENDIX XX* to assist others in future attempts to achieve better results.  However, the line between false and negative positives may be blurred in some of these instances and warrant further work, for example **Ericaceae pollen grains were observed in a number of samples in trace quantities...**.

@@

  The remarkably clear foraging preferences of *Bombus*, both here and across a great many localities globally, reiterates the needs for land managers to maintain relatively high amounts of members of the Fabaceae, Boraginaceae, and Ranunculaceae across the landscape. Numerous historic, and some ongoing, land management actions reduce the ability of many landscapes to support stable populations of *Bombus*. Historic, and occasionally current, livestock grazing may be associated with the targeted removal of many species of plants which are known to have compounds toxic to cattle. In particular, the removal of locoweeds (Fabaceae: *Astragalus* L. & *Oxytropis* DC.) and larkspurs (Ranunculaeae: *Delphinium*) were common across public lands administered by the United States Forest Service, and to an extent the Bureau of Land Management (@ralphs1988herbicide, @aldous1919eradicating, @ralphs2003mechanism). Further actions, generally initiated by settlers to gain water rights or to support mining operations, involved the channelization and incising of streams, and culling of beavers, reducing the extent of wetlands and associated mesic meadows which provide habitat for many species of *Mertensia* (Boraginaceae) widely distributed across Western North America (). Fire suppression further resulted in the succession of many Aspen groves to Conifer stands, decreasing the mosaic of age structured habitats in many landscapes, adversely effects habitat for tall *Mertensia* species. Finally the effects of Nitrogen deposition, especially given the West's growing population still pose adverse effects on the abundance of a variety of species of Fabaceae (see @stevens2018atmospheric). Current solutions to these issues, involve targeted burns, reintroduction of beavers and beaver habitats, and the possibility of re-seeding a variety of 'locoweeds' and 'larkspurs' in areas now seldom used for grazing. The highly enthusiastic response of land managers, and homeowners, to plant *Ascelpias* L. to improve Monarch Butterfly (*Danaus plexippus* L.) habitat provides an effective framework for the latter.

@@

Appendix 1 - Site Maps

@@

Appendix 2 - Species Distribution Models Predictors

Layer   LM                     Description                            Source
----   ----    ------------------------------------------     ---------------------
1.                   Mean annual cloudiness - MODIS            Wilson et al. 2016
2.                 Cloudiness seasonality 1 - MODIS            Wilson et al. 2016 
3.                 Cloudiness seasonality 2 - MODIS            Wilson et al. 2016 
4.                 Cloudiness seasonality 3 - MODIS            Wilson et al. 2016
5.                Beginning of the frost-free period               Wang et al.
6.                   Climatic moisture deficit                     Wang et al.
7.                   Degree-days above 5C from                     Wang et al.
8.                   Mean annual precipitation                     Wang et al.
9.                Mean annual precipitation as snow                Wang et al.
10.                   Temperature seasonality                      Wang et al.
11.             2015 Percent Grass/Herbaceous cover - MODIS         (MOD44B)
12.            2015 Percent Tree cover from Landsat 7/8              (GLCF)
13.             Soil probability of bedrock (R Horizon)             SoilGrids
14.               Soil organic carbon (Tonnes / ha)                 SoilGrids
15.                    Surface soil pH in H2O                       SoilGrids
16.                   Surface soil percent sand                     SoilGrids
17.                         Soil USDA class                         SoilGrids
18.                       Topographic elevation                    EarthEnv DEM
19.                 Topographic elevation, moving window.          EarthEnv DEM
20.                   Topographic percent slope                    EarthEnv DEM 
21.                   Topographic wetness index                    EarthEnv DEM 
22.                    Topographic aspect from                     EarthEnv DEM
23.             Annual potential solar radiation computed             r.sun
24.            Estimated actual (w/-cloud) solar radiation  r.sun / Wilson et al. 2016
25.             Log-transformed distance to surface water   Global Surface Water Explorer
26.                   Percent surface water                 Global Surface Water Explorer 
----        -------------------------------------------      ---------------------

@@

Appendix 3 - Molecular Reference Specimen Table

Appendix 4 - All Pollen Reference Slides Used to Establish Morphotypes

### POLLEN CLUSTER RESULTS SHOULD BE HERE

@@

@@

Appendix XX - Comparison of Kraken2, Bracken, and BLAST

@@

Random Forest have high bias and low variance, where boosted regressions trees have low bias and high variances.
